Linting Linearity in Core/System FC

### 2023-01-12 report.tex
minor changes

% BNF grammars
\usepackage[center]{formal-grammar}

@@

% Linear types are cool
% Among the few, Haskell has linear types
% But Haskell differs from other languages with linear types
%   * added 31 years after inception
%   * added to its core
% Core has linear types for good reasons
% But they are broken -- why?
%   * Optimizations changes source -- e.g. if recursive lets are introduced, how to type linearity?
%

Linear type systems\cite{} add expressiveness to existing type systems by
stating that linear resources must be used \emph{exactly once}. In programming
languages with a linear type system, not using certain resources or using them
twice are type errors. Linear types can thus be used to, for example, guarantee
that file handles, socket descriptors, and allocated memory are freed exactly
once (leaks and double-frees become type errors)\cite{}, implement safe
high-performance language interoperability\cite{}, guarantee that quantum
entangled states are not duplicated~\cite{}, handle mutable state safely\cite{},
something on session types guarantees\cite{}, and much more.

@@

Core was imbued with linear types because an implementation detail required it,
and, more importantly, to guarantee linear programs remain linear after being
desugared and optimized: much the same way a \emph{typed} Core ensures that
desugaring, and optimizing transformations, are correctly implemented and result
in a \emph{well-typed} program -- a design decision that has well served the
compiler developers with a safety net for a long time. It is crucial that a
program behaviour enforced by linear types, regarding the usage of certain
resources, is \emph{not} changed in the desugaring or optimization process
(optimizations should not destroy linearity!) and a linearity aware Core
typechecker can guarantee that.

@@

\begin{grammar}
    \firstcase{\t}{\star}{Unit type}
    \otherform{t \tensor t'}{Multiplicative conjunction type}
    % t ::= a $\lolli$ b
\end{grammar}

>>>>

@@

% Statically typed languages 
% Linear types are cool
% Among the few, Haskell has linear types
% But Haskell differs from other languages with linear types
%   * added 31 years after inception
%   * added to its core
% Core has linear types for good reasons
% But they are broken -- why?
%   * Optimizations changes source -- e.g. if recursive lets are introduced, how to type linearity?
% Goals - O que eu vou fazer
% Document Structure (pode-se eventualmente combinar com a seccao dos goals)

%\section{Introduction}

Linear type systems~\cite{} add expressiveness to existing type systems by
stating that linear resources must be used \emph{exactly once}. In programming
languages with a linear type system, not using certain resources or using them
twice are type errors. Linear types can thus be used to, for example, guarantee
that file handles, socket descriptors, and allocated memory are freed exactly
once (leaks and double-frees become type errors)~\cite{}, implement safe
high-performance language interoperability~\cite{}, guarantee that quantum
entangled states are not duplicated~\cite{}, handle mutable state safely~\cite{},
something on session types guarantees~\cite{}, and much more.

@@

Core was imbued with linear types to guarantee linear programs remain linear after being
desugared and optimized: much the same way a \emph{typed} Core ensures that
desugaring, and optimizing transformations, are correctly implemented and result
in a \emph{well-typed} program -- a design decision that has well served the
compiler developers with a safety net for a long time. It is crucial that a
program behaviour enforced by linear types, regarding the usage of certain
resources, is \emph{not} changed in the desugaring or optimization process
(optimizations should not destroy linearity!) and a linearity aware Core
typechecker can guarantee that.

@@

\[
  \begin{array}{lcl}
    s,t & ::= & \star\\
        & \mid & t \lolli s\\
       \dots
   \end{array}
\]
